# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

1. Load the data (i.e. `read.csv()`)
2. Process/transform the data (if necessary) into a format suitable
   for your analysis

   ```{r, echo="true"}
   data <- read.csv(unz("activity.zip", "activity.csv"),
                    header=TRUE,
                    colClasses=c("numeric", "Date", "numeric"))
   ```

## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in
the dataset.

1. Make a histogram of the total number of steps taken each day

   The following histogram shows the distribution of the total number
of steps per day over the number of days that total was
reached.

   Originally I used `na.rm=TRUE` here, so missing measurements were
not included as part of the sum, and days without any measurements had
a sum of zero. However this made the "Imputing Values" section of the
assignment less interesting, so although I believe it was a superior
method I have changed it so missing values taint the daily steps
aggregation.

   ```{r, echo="true"}
   dailySteps <- rowsum(data$steps, data$date, na.rm=FALSE)
   hist(dailySteps, breaks=10, main="Histogram of daily steps")
   meanSteps <- mean(dailySteps)
   ```

2. Calculate and report the **mean** and **median** total number of
   steps taken per day

   Answer: The daily step average is `r meanSteps`. Here it is shown
along with the distribution of the total number of steps per day
across the whole data set, including the mean and the median:

   ```{r, echo="true"}
   boxplot(dailySteps, main="Daily steps")
   abline(h=meanSteps)
   legend("topright", summary(dailySteps))
   ```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. `type = "l"`) of the 5-minute
   interval (x-axis) and the average number of steps taken, averaged
   across all days (y-axis)

   ```{r, echo="true"}
   meanStepsByInterval <- aggregate(steps ~ interval, data=data, "mean")
   plot(meanStepsByInterval, type="l")
   ```

2. Which 5-minute interval, on average across all the days in the
   dataset, contains the maximum number of steps?

   ```{r, echo="true"}
   maxRow <- which.max(meanStepsByInterval$steps)
   maxInterval <- meanStepsByInterval[maxRow,]$interval
   ```

   Answer: Interval `r maxInterval` has the highest number of steps,
averaged across all the days in the dataset.

## Imputing missing values

Note that there are a number of days/intervals where there are missing
values (coded as `NA`). The presence of missing days may introduce
bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the
dataset (i.e. the total number of rows with `NA`s)

   ```{r, echo="true"}
   missingRowCount <- sum(!complete.cases(data))
   ```

   Answer: There are `r missingRowCount` rows with missing values in
the dataset.

2. Devise a strategy for filling in all of the missing values in the
   dataset. The strategy does not need to be sophisticated. For
   example, you could use the mean/median for that day, or the mean
   for that 5-minute interval, etc.

   Answer: Using the mean for the day (this seems like a bad idea, as
the lecturer stated, but I'm doing it as it's a required part of this
assignment.)

   ```{r, echo="true"}
   meanForInterval <- tapply(data$steps, data$interval, "mean", na.rm=TRUE)
   hist(meanForInterval)

   library(stringr)
   stepsOrMeanForInterval <- function(row) {
           if (is.na(row[1])) {
                   interval <- str_trim(row[3])
                   meanForInterval[[interval]]
           }
           else {
                   as.numeric(row[1])
           }
   }
   ```

3. Create a new dataset that is equal to the original dataset but with
   the missing data filled in.

   ```{r, echo="true"}
   imputedSteps <- apply(data, 1, stepsOrMeanForInterval)
   newdata <- within(data, { steps <- imputedSteps })
   ```

4. Make a histogram of the total number of steps taken each day and
   calculate and report the mean and median total number of steps
   taken per day. Do these values differ from the estimates from the
   first part of the assignment? What is the impact of imputing
   missing data on the estimates of the total daily number of steps?

   ```{r, echo="true"}
   newDailySteps <- rowsum(newdata$steps, newdata$date, na.rm=FALSE)
   hist(newDailySteps, breaks=10, main="Histogram of daily steps (imputed)")
   newMeanSteps <- mean(newDailySteps)

   boxplot(dailySteps[,1], newDailySteps[,1],
           main="Daily steps (original vs imputed)")
   legend("topleft", summary(dailySteps), title="Original")
   legend("topright", summary(newDailySteps), title="Imputed")
   ```

   Answer: Do these values differ? Not really, though from the boxplot
we can see the imputed values result in a "tight" pattern, with less
variance from the mean.

## Are there differences in activity patterns between weekdays and weekends?

For this part the `weekdays()` function may be of some help here. Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels -
   "weekday" and "weekend" indicating whether a given date is a
   weekday or weekend day.

   ```{r, echo="true"}
   dayLevels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday",
                  "Saturday", "Sunday")
   data$dayOfWeek <- factor(weekdays(data$date), levels=dayLevels, ordered=TRUE)
   data$dayType <- factor(data$dayOfWeek >= "Monday" & data$dayOfWeek <= "Friday",
                          labels=c("weekend", "weekday"))
   ```

2. Make a panel plot containing a time series plot (i.e. `type = "l"`)
   of the 5-minute interval (x-axis) and the average number of steps
   taken, averaged across all weekday days or weekend days
   (y-axis). The plot should look something like the following, which
   was creating using simulated data:

   ```{r, echo="true"}
   stepsByDayType <- aggregate(steps ~ interval * dayType, data=data, mean)
   xyplot(steps ~ interval | dayType, data=stepsByDayType,
          type="l", layout=c(1,2))
   ```
